---
title: "Intro2MordernStats - Day1"
author: "Wenbin Guo"
date: "2025-02-11"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
if (!requireNamespace("DescTools", quietly = TRUE)) {
  install.packages("DescTools")
}
```

# Introduction
This is introduction to modern statistics Day1. We will cover basic concepts, probability, distributions, and descriptive statistics.


# Probability
```{r}
# Monte Carlo Simulation to Estimate Pi
set.seed(123)  # For reproducibility

n <- 1000  # Number of random points
x <- runif(n)  # Generate n random x-coordinates between 0 and 1
y <- runif(n)  # Generate n random y-coordinates between 0 and 1

# Check if the points are inside the unit circle (radius = 1)
inside_circle <- (x^2 + y^2) <= 1

# Estimate Pi
pi_estimate <- 4 * sum(inside_circle) / n
cat("Estimated Pi:", pi_estimate, "\n")

# Plot the points
plot(x, y, col = ifelse(inside_circle, "red", "blue"), pch = 20, asp = 1,
     main = bquote(n == .(n) ~ "; " ~ pi %~~% .(round(pi_estimate, 4))),
     xlab = "", ylab = "",
     xlim = c(0,1), ylim = c(0,1))
```

TODO: plot how the estimate of Pi converges as n increases, note you should be able to 
```{r}

```



# Distributions
## discrete distributions
### uniform
```{r}
# Uniform Distribution (a = 1, b = 6)
x <- 1:6
y <- rep(1/6, 6)

plot(x, y, type = "h", lwd = 2, col = "blue", main = "PMF of Uniform Distribution", 
     xlab = "x", ylab = "P(X = x)")
points(x, y, pch = 16, col = "blue")
```

### bernoulli
```{r}
# Bernoulli Distribution (p = 0.3)
x <- c(0, 1)
y <- dbinom(x, size = 1, prob = 0.3)

plot(x, y, type = "h", lwd = 2, col = "blue", main = "PMF of Bernoulli Distribution", 
     xlab = "x", ylab = "P(X = x)", xlim = c(-0.5, 1.5))
points(x, y, pch = 16, col = "blue")
```

### binomial
```{r}
# Binomial Distribution (n = 10, p = 0.5)
x <- 0:10
y <- dbinom(x, size = 10, prob = 0.5)

plot(x, y, type = "h", lwd = 2, col = "blue", main = "PMF of Binomial Distribution", 
     xlab = "x", ylab = "P(X = x)")
points(x, y, pch = 16, col = "blue")
```

### negative binomial
```{r}
# Negative Binomial Distribution (r = 5, p = 0.4)
x <- 0:15
y <- dnbinom(x, size = 5, prob = 0.4)

plot(x, y, type = "h", lwd = 2, col = "blue", main = "PMF of Negative Binomial Distribution", 
     xlab = "x", ylab = "P(X = x)")
points(x, y, pch = 16, col = "blue")
```


### geometric
```{r}
# Geometric Distribution (p = 0.3)
x <- 1:15
y <- dgeom(x - 1, prob = 0.3)

plot(x, y, type = "h", lwd = 2, col = "blue", main = "PMF of Geometric Distribution", 
     xlab = "x", ylab = "P(X = x)")
points(x, y, pch = 16, col = "blue")
```

### hypergeometric
```{r}
# Hypergeometric Distribution (N = 20, K = 10, n = 5)
x <- 0:5
y <- dhyper(x, m = 10, n = 10, k = 5)

plot(x, y, type = "h", lwd = 2, col = "blue", main = "PMF of Hypergeometric Distribution", 
     xlab = "x", ylab = "P(X = x)")
points(x, y, pch = 16, col = "blue")
```


### poisson
```{r}
# Poisson Distribution (lambda = 3)
x <- 0:15
y <- dpois(x, lambda = 3)

plot(x, y, type = "h", lwd = 2, col = "blue", main = "PMF of Poisson Distribution", 
     xlab = "x", ylab = "P(X = x)")
points(x, y, pch = 16, col = "blue")
```

## continuous
### uniform
```{r}
# Continuous Uniform Distribution (a = 0, b = 1)
x <- seq(0, 1, length.out = 100)
y <- dunif(x, min = 0, max = 1)

plot(x, y, type = "l", lwd = 2, col = "blue", main = "PDF of Continuous Uniform Distribution", 
     xlab = "x", ylab = "f(x)")
```


### normal
```{r}
# Normal Distribution (mean = 0, sd = 1)
x <- seq(-4, 4, length.out = 200)
y <- dnorm(x, mean = 0, sd = 1)

plot(x, y, type = "l", lwd = 2, col = "blue", main = "PDF of Normal Distribution", 
     xlab = "x", ylab = "f(x)")
```


### t-distribution
```{r}
# t-Distribution (df = 5)
x <- seq(-5, 5, length.out = 200)
y <- dt(x, df = 5)

plot(x, y, type = "l", lwd = 2, col = "blue", main = "PDF of t-Distribution", 
     xlab = "x", ylab = "f(x)")
```

TODO:
plot the $t$-distribution with different degrees of freedom (df = 1, 2, 5, 10, 20, 30, 50, 100) on the same plot
```{r}

```

plot the $t$-distribution along with the standard normal distribution on the same plot, what did you find?
```{r}

```



### chi-square
```{r}
# Chi-Square Distribution (df = 3)
x <- seq(0, 10, length.out = 200)
y <- dchisq(x, df = 3)

plot(x, y, type = "l", lwd = 2, col = "blue", main = "PDF of Chi-Square Distribution", 
     xlab = "x", ylab = "f(x)")
```

### F-distribution
```{r}
# F-Distribution (d1 = 5, d2 = 10)
x <- seq(0, 5, length.out = 200)
y <- df(x, df1 = 5, df2 = 10)

plot(x, y, type = "l", lwd = 2, col = "blue", main = "PDF of F-Distribution", 
     xlab = "x", ylab = "f(x)")
```


### exponential
```{r}
# Exponential Distribution (rate = 1)
x <- seq(0, 5, length.out = 200)
y <- dexp(x, rate = 1)

plot(x, y, type = "l", lwd = 2, col = "blue", main = "PDF of Exponential Distribution", 
     xlab = "x", ylab = "f(x)")
```


### beta
```{r}
# Beta Distribution (shape1 = 2, shape2 = 5)
x <- seq(0, 1, length.out = 200)
y <- dbeta(x, shape1 = 2, shape2 = 5)

plot(x, y, type = "l", lwd = 2, col = "blue", main = "PDF of Beta Distribution", 
     xlab = "x", ylab = "f(x)")

```



# Descriptive Statistics
## Measures of Central Tendency
```{r}
my_vec = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 5)

mean(my_vec)  # Calculate the mean of a vector
median(my_vec)  # Calculate the median of a vector


library(DescTools)
Mode(my_vec)  # Calculate the mode of a vector
```

## Measures of Dispersion
```{r}
var(my_vec)  # Calculate the variance of a vector
sd(my_vec)   # Calculate the standard deviation of a vector
range(my_vec)# Calculate the range of a vector
quantile(my_vec)  # Calculate the quartiles of a vector
IQR(my_vec)  # Calculate the interquartile range of a vector
```

## visualize distribution
### histogram
```{r}
hist(my_vec)  # Create a histogram of a vector
```

### boxplot
```{r}
boxplot(my_vec)  # Create a boxplot of a vector
```


