---
title: "Introduction to Modern Statistics - Day2"
author: "Wenbin Guo"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```


# Introduciton
This is introduction to modern statistics Day2. We will cover hypothesis testing, permutation test, bootstrap, multiple testing, etc.



# Population & samples
## population
```{r}
# Load required library
library(ggplot2)

# Generate data for a standard normal distribution
x_vals <- seq(-4, 4, by = 0.01)
pdf_vals <- dnorm(x_vals, mean = 0, sd = 1)

# Create the plot
ggplot(data.frame(x = x_vals, y = pdf_vals), aes(x = x, y = y)) +
    geom_line(color = "blue", size = 1) +
    labs(title = "Probability Density Function of Standard Gaussian", x = "x", y = "Density") +
    annotate("text", x = 3, y = 0.35, label = "Mean = 0", color = "red", size = 5) +
    annotate("text", x = 3, y = 0.3, label = "Variance = 1", color = "red", size = 5) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "darkred", size = 0.8) +
    theme_minimal(base_size = 14)

```

## samples
```{r}
set.seed(42)  # For reproducibility
n_samples <- 100
samples <- rnorm(n_samples, mean = 0, sd = 1)
sample_mean <- mean(samples)
sample_variance <- var(samples)

# Create the histogram with sample mean and variance annotations
ggplot(data.frame(x = samples), aes(x = x)) +
  geom_histogram(aes(y = ..density..), bins = 15, fill = "skyblue", color = "black", alpha = 0.7) +
  #stat_function(fun = dnorm, args = list(mean = 0, sd = 1), color = "blue", size = 1) +
  geom_vline(xintercept = sample_mean, linetype = "dashed", color = "red", size = 0.8) +
  annotate("text", x = sample_mean + 0.3, y = 0.35, label = paste0("Sample Mean = ", round(sample_mean, 2)), color = "red", size = 5, hjust = 0) +
  annotate("text", x = sample_mean + 0.3, y = 0.3, label = paste0("Sample Variance = ", round(sample_variance, 2)), color = "red", size = 5, hjust = 0) +
  labs(
    title = "Histogram of 100 Samples from Standard Gaussian",
    x = "x",
    y = "Density"
  ) +
  theme_minimal(base_size = 14)
```


TODO: change the random seed, or change the number of samples, see how the histogram vary


# Theorem examples
## Law of Large Numbers
```{r}
# Law of Large Numbers (LLN)
set.seed(123)  # For reproducibility
n <- 100000  # Number of simulations
rolls <- sample(1:6, size = n, replace = TRUE)  # Simulate rolling a die
sample_means <- cumsum(rolls) / (1:n)  # Compute cumulative mean

# Plot the convergence of the sample mean
plot(1:n, sample_means, type = "l", col = "blue", 
     main = "Law of Large Numbers",
     xlab = "Number of Rolls", ylab = "Sample Mean")
abline(h = 3.5, col = "red", lwd = 2, lty = 2)  # Population mean (expected value)
legend("topright", legend = "Population Mean = 3.5", col = "red", lty = 2)
```

## Central Limit Theorem
```{r}
set.seed(123)  # For reproducibility

# Parameters
n <- 30            # Sample size for each sample
num_samples <- 5000  # Number of samples
dist_function <- function(n) rexp(n, rate = 1)  # Change this to your desired distribution

# Generate sample means
sample_means <- replicate(num_samples, mean(dist_function(n)))

# Plot the histogram of the sample means
hist(sample_means, breaks = 40, prob = TRUE, col = "lightblue", 
     main = "Central Limit Theorem",
     xlab = "Sample Means")

# Overlay normal distribution curve
curve(dnorm(x, mean = mean(sample_means), sd = sd(sample_means)), 
      col = "red", lwd = 2, add = TRUE)

legend("topright", legend = "Normal Distribution", col = "red", lty = 1)
```

TODO: change the data generating distribution, see how the histogram change?


# Hypothesis testing
## example
```{r}
# Load library
library(ggplot2)

# Create a sequence of values for x-axis
x_vals <- seq(-4, 4, length.out = 1000)

# Calculate PDF for two normal distributions
pdf1 <- dnorm(x_vals, mean = 0, sd = 1)      # Group 1: N(0, 1)
pdf2 <- dnorm(x_vals, mean = 0.2, sd = 1)    # Group 2: N(0.2, 1)

# Create a data frame for plotting
pdf_data <- data.frame(
  x = rep(x_vals, 2),
  y = c(pdf1, pdf2),
  group = rep(c("Group 1", "Group 2"), each = length(x_vals))
)

# Plot density curves
ggplot(pdf_data, aes(x = x, y = y, color = group)) +
  geom_line(size = 1) +
  labs(
    title = "PDF of Two Gaussian Distributions",
    x = "x",
    y = "Density",
    caption = "Null hypothesis: Same mean\nAlternative hypothesis: Different means"
  ) +
  scale_color_manual(values = c("blue", "red")) +
  theme_minimal()
```

```{r}
# Load library
library(ggplot2)

# Generate 50 samples from each group
set.seed(42)
samples_group1 <- rnorm(50, mean = 0, sd = 1)      # Group 1: N(0, 1)
samples_group2 <- rnorm(50, mean = 0.2, sd = 1)    # Group 2: N(0.2, 1)

# Combine into a data frame
sample_data <- data.frame(
  value = c(samples_group1, samples_group2),
  group = rep(c("Group 1", "Group 2"), each = 50)
)

# Calculate sample means for annotation
mean_group1 <- mean(samples_group1)
mean_group2 <- mean(samples_group2)

# Plot histogram
ggplot(sample_data, aes(x = value, fill = group)) +
  geom_histogram(aes(y = ..density..), bins = 15, position = "identity", alpha = 0.5) +
  geom_vline(aes(xintercept = mean_group1), color = "blue", linetype = "dashed", size = 1, show.legend = FALSE) +
  geom_vline(aes(xintercept = mean_group2), color = "red", linetype = "dashed", size = 1, show.legend = FALSE) +
  labs(
    title = "Histogram of 50 Samples from Two Gaussian Distributions",
    x = "Value",
    y = "Density",
    caption = "Dashed lines indicate sample means for Group 1 (blue) and Group 2 (red)"
  ) +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal()
```

```{r}
t.test(samples_group1, samples_group2)
```




## Parametric tests
### One-sample z-test
Goal: Test if the sample mean is equal to a specified value when the population variance is known.
```{r}
library(BSDA)
set.seed(123)
data_one_sample <- rnorm(100, mean = 50, sd = 10)  # Generate sample data
z.test(data_one_sample, mu = 50, sigma.x = 10)  # Test if mean is equal to 50
```

### Two-sample z-test
Goal: Test if two independent samples have the same population mean.
```{r}
set.seed(123)
group1 <- rnorm(100, mean = 50, sd = 10)
group2 <- rnorm(100, mean = 52, sd = 10)

z.test(group1, group2, sigma.x = 10, sigma.y = 10)  # Test if the means are equal
```


### One-sample t-test
Goal: Test whether the mean of a single sample is equal to a specified value (e.g., mean = 5).
```{r}
set.seed(123)
data_one_sample <- rnorm(30, mean = 5, sd = 2)  # Generate 30 random values with mean 5 and sd 2
t.test(data_one_sample, mu = 5)  # Test if the mean is equal to 5
```

### Two-sample t-test
Goal: Compare the means of two independent samples.
```{r}
set.seed(123)
group1 <- rnorm(30, mean = 5, sd = 2)
group2 <- rnorm(30, mean = 6, sd = 2)
t.test(group1, group2) 
```

```{r}
set.seed(123)
group1 <- rnorm(30, mean = 5, sd = 2)
group2 <- rnorm(30, mean = 6, sd = 2)
t.test(group1, group2, alternative = "less") 
```


### Paired t-test
Goal: Compare the means of two related samples (before and after measurements).
```{r}
set.seed(123)
before <- rnorm(30, mean = 5, sd = 2)
after <- before + rnorm(30, mean = 0.5, sd = 1)
t.test(before, after, paired = TRUE)  # Test for differences between paired samples
```

### ANOVA
Goal: Test for differences in means across multiple groups.
```{r}
set.seed(123)
group <- factor(rep(1:3, each = 10))
values <- c(rnorm(10, mean = 5), rnorm(10, mean = 6), rnorm(10, mean = 7))
anova_result <- aov(values ~ group)
summary(anova_result)
```


### F-test
Goal: Compare the variances of two independent samples. Useful for checking the assumption of equal variances in a two-sample t-test.
```{r}
set.seed(123)
group1 <- rnorm(30, mean = 5, sd = 1)
group2 <- rnorm(30, mean = 5, sd = 3)

var.test(group1, group2)  # Test for equal variances
```


### Chi-square test
Goal: Test for the independence of two categorical variables.
```{r}
set.seed(123)
observed <- matrix(c(50, 30, 20, 40), nrow = 2)  # Create a 2x2 contingency table
print(observed)
chisq.test(observed)  # Test for independence between rows and columns
```


## Non-parametric tests
### Mann-Whitney U test
Goal: Compare two independent samples without assuming normality.  The null hypothesis is that the distributions of the two groups are identical. It’s a non-parametric alternative to the two-sample t-test.
```{r}
set.seed(123)
group1 <- rnorm(30, mean = 5, sd = 2)
group2 <- rnorm(30, mean = 6, sd = 2)
wilcox.test(group1, group2)
```

### Kruskal-Wallis test
Goal: Test for differences in medians across multiple groups without assuming normality.
```{r}
set.seed(123)
group <- factor(rep(1:3, each = 10))
values <- c(rnorm(10, mean = 5), rnorm(10, mean = 6), rnorm(10, mean = 7))
kruskal.test(values ~ group)
```


### Kolmogorov-Smirnov Test
Goal: Compare a sample with a reference distribution or compare two samples. Useful for checking if two distributions differ significantly.
```{r}
set.seed(123)
sample1 <- rnorm(50)
sample2 <- rnorm(50, mean = 1)
ks.test(sample1, sample2)  # Two-sample test
```

## Other commonly-used test
###  Shapiro-Wilk Test
Goal: Test if a sample comes from a normal distribution. Used to check the normality assumption before applying parametric tests like the t-test or ANOVA.
```{r}
set.seed(123)
data_normal <- rnorm(50)
shapiro.test(data_normal)
```




## p-value distribution
### histogram
```{r}
p_vec = runif(1000)

hist(p_vec, breaks = 20, main = "Histogram of p-values", xlab = "p-value", col = "skyblue")
```

### QQ plot
```{r}
library(qqman)

qq(p_vec, main = "QQ Plot of p-values", col = "skyblue")
```


# Permutation test and bootstrap
## Permutation test
```{r}
library(openintro)

data("sex_discrimination")
table(sex_discrimination$sex, sex_discrimination$decision)
```

```{r}
# Calculate the observed difference in promotion rates
obs_diff <- sex_discrimination %>%
  group_by(rev(sex)) %>% 
  summarise(promotion_rate = mean(decision == "promoted")) %>%
  summarise(diff = diff(promotion_rate)) %>%
  pull(diff)

# Permutation test
set.seed(42)  # for reproducibility
n_perm <- 1000
perm_diffs <- replicate(n_perm, {
  perm_data <- sex_discrimination %>%
    mutate(decision = sample(decision))  # Permute the decision variable
  perm_diff <- perm_data %>%
    group_by(rev(sex)) %>%
    summarise(promotion_rate = mean(decision == "promoted")) %>%
    summarise(diff = diff(promotion_rate)) %>%
    pull(diff)
  perm_diff
})

# Calculate p-value
p_value <- mean(abs(perm_diffs) >= abs(obs_diff))

# Visualize the permutation distribution
hist(perm_diffs, breaks = 50, main = "Permutation Test Distribution",
     xlab = "Difference in Promotion Rates", border = "blue")
abline(v = obs_diff, col = "red", lwd = 2, lty = 2)
text(obs_diff, max(table(cut(perm_diffs, breaks = 50))), 
     paste("Observed Diff =", round(obs_diff, 3)), pos = 4, col = "red")

# Print observed difference and p-value
cat("Observed Difference in Promotion Rates:", obs_diff, "\n")
cat("P-value:", p_value, "\n")
```



## Bootstrap
```{r}
data("cpr")

table(cpr$group, cpr$outcome)
```

```{r}
# Calculate observed survival rates for each group
observed_diff <- cpr %>%
  group_by(group) %>%
  summarise(survival_rate = mean(outcome == "survived")) %>%
  summarise(diff = diff(survival_rate)) %>%  # difference: treatment - control
  pull(diff)

# Bootstrap function
set.seed(123)  # for reproducibility
n_boot <- 1000
bootstrap_diffs <- replicate(n_boot, {
  boot_data <- cpr %>%
    sample_frac(replace = TRUE)  # Resample with replacement
  boot_diff <- boot_data %>%
    group_by(group) %>%
    summarise(survival_rate = mean(outcome == "survived")) %>%
    summarise(diff = diff(survival_rate)) %>%
    pull(diff)
  boot_diff
})

# Calculate 95% confidence interval (percentile method)
ci_lower <- quantile(bootstrap_diffs, 0.025)
ci_upper <- quantile(bootstrap_diffs, 0.975)

# Print results
cat("Observed Difference in Survival Rates:", observed_diff, "\n")
cat("95% Confidence Interval: [", ci_lower, ",", ci_upper, "]\n")

# Visualize the bootstrap distribution
hist(bootstrap_diffs, breaks = 50, main = "Bootstrap Distribution of Survival Rate Difference",
     xlab = "Difference in Survival Rates (Treatment - Control)", border = "blue")
abline(v = c(ci_lower, ci_upper), col = "red", lwd = 2, lty = 2)
abline(v = observed_diff, col = "darkgreen", lwd = 2)
text(observed_diff, max(table(cut(bootstrap_diffs, breaks = 50))), 
     paste("Observed Diff =", round(observed_diff, 3)), pos = 4, col = "darkgreen")
```



# Multiple testing
```{r}
# Set seed for reproducibility
set.seed(42)

# Generate synthetic data
n_genes <- 1000
n_samples <- 10
n_control <- n_samples / 2
n_disease <- n_samples / 2

# Generate control data
data_control <- matrix(rnorm(n_genes * n_control, mean = 5, sd = 1), nrow = n_genes)

# Copy control data for disease group and add an effect to 10% of genes
n_diff_genes <- ceiling(0.10 * n_genes)
data_disease <- data_control
data_disease[1:n_diff_genes, ] <- data_disease[1:n_diff_genes, ] + 3  # Effect size


# Combine into one dataset
data <- cbind(data_control, data_disease)
colnames(data) <- paste0("Sample_", 1:n_samples)
rownames(data) <- paste0("Gene_", 1:n_genes)
true_diff_genes = rownames(data)[1:n_diff_genes]

# Perform t-tests for each gene
pvalues <- apply(data, 1, function(row) {
  t.test(row[1:n_control], row[(n_control + 1):n_samples], var.equal = FALSE)$p.value
})
```


## Bonferroni correction
```{r}
bonferroni_corrected <- p.adjust(pvalues, method = "bonferroni")
head(bonferroni_corrected)
```


## False discovery rate
```{r}
fdr_corrected <- p.adjust(pvalues, method = "fdr")
head(fdr_corrected)
```

## compare
```{r}
bonforoni_sig_genes = rownames(data)[bonferroni_corrected < 0.05]
fdr_sig_genes = rownames(data)[fdr_corrected < 0.05]
```

```{r}
bonforoni_sig_genes
length(bonforoni_sig_genes)
```

```{r}
fdr_sig_genes
length(fdr_sig_genes)
```

